---
title: "Discover and download datasets and files from the cellxgene data portal"
author:
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center
  email: Martin.Morgan@RoswellPark.org
package: cellxgenedp
output:
    BiocStyle::html_document
abstract: |
    The cellxgene data portal (https://cellxgene.cziscience.com/) provides a
    graphical user interface to collections of single-cell sequence data 
    processed in standard ways to 'count matrix' summaries. The cellxgenedp
    package provides an alternative, R-based inteface, allowing flexible data
    discovery, viewing, and downloading.
vignette: |
    %\VignetteIndexEntry{Discover and download datasets and files from the cellxgene data portal}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

# Installation and use

This package is not yet available on _Bioconductor_. Install from
GitHub using the following commands

```{r, eval = FALSE}
installed_packages <- rownames(installed.packages())

## packages required for installation
pkgs <- c("remotes", "BiocManager")
required_pkgs <- pkgs[!pkgs %in% installed_packages]
install.packages(required_pkgs, repos = "https://cloud.r-project.org")

## the cellxgenedp package, and packages used in the vignette
vignette_dependencies <- c("zellkonverter", "SingleCellExperiment")
required_pkgs <-
    vignette_dependencies[!vignette_dependencies %in% installed_packages]
BiocManager::install(c(required_pkgs, "mtmorgan/cellxgenedp"))
```

Load the package into your current _R_ session. We make extensive use
of the dplyr packages, and at the end of the vignette use
SingleCellExperiment and zellkonverter, so load those as well.

```{r}
suppressPackageStartupMessages({
    library(zellkonverter)
    library(SingleCellExperiment) # load early to avoid masking dplyr::count()
    library(dplyr)
    library(cellxgenedp)
})
```

# Collections, datasets and files

Retrieve metadata about resources available at the cellxgene data
portal using `db()`:

```{r}
db <- db()
```

Printing the `db` object provides a brief overview of the available
data, as well as hints, in the form of functions like `collections()`,
for further exploration.

```{r}
db
```

The portal organizes data hierarchically, with 'collections'
(research studies, approximately), 'datasets', and 'files'. Discover
data using the corresponding functions.

```{r}
collections(db)

datasets(db)

files(db)
```

Each of these resources has a unique primary identifier (e.g.,
`file_id`) as well as an identifier describing the relationship of the
resource to other components of the database (e.g.,
`dataset_id`). These identifiers can be used to 'join' information
across tables.

## Using `dplyr` to navigate data

A collection may have several datasets, and datasets may have several
files. For instance, here is the collection with the most datasets

```{r}
collection_with_most_datasets <-
    datasets(db) |>
    count(collection_id, sort = TRUE) |>
    slice(1)
```

We can find out about this collection by joining with the
`collections()` table.

```{r}
left_join(
    collection_with_most_datasets |> select(collection_id),
    collections(db),
    by = "collection_id"
) |> glimpse()
```

We can take a similar strategy to identify all datasets belonging to
this collection
```{r}
left_join(
    collection_with_most_datasets |> select(collection_id),
    datasets(db),
    by = "collection_id"
)
```

## `facets()` provides information on 'levels' present in specific columns

Notice that some columns are 'lists' rather than atomic vectors like
'character' or 'integer'.

```{r}
datasets(db) |>
    select(where(is.list))
```

This indicates that at least some of the datasets had more than one
type of `assay`, `cell_type`, etc. The `facets()` function provides a
convenient way of discovering possible levels of each column, e.g.,
`assay`, `organism`, `ethnicity`, or `sex`, and the number of datasets with
each label.

```{r}
facets(db, "assay")
facets(db, "ethnicity")
facets(db, "sex")
```

## Filtering faceted columns

Suppose we were interested in finding datasets from the 10x 3' v3
assay (`ontology_term_id` of `EFO:0009922`) containing individuals of
African American ethnicity, and female sex. Use the `facets_filter()`
utility function to filter data sets as needed

```{r}
african_american_female <-
    datasets(db) |>
    filter(
        facets_filter(assay, "ontology_term_id", "EFO:0009922"),
        facets_filter(ethnicity, "label", "African American"),
        facets_filter(sex, "label", "female")
    )
```

There are `r nrow(african_american_female)` datasets satisfying our
criteria. It looks like there are up to

```{r}
african_american_female |>
    summarise(total_cell_count = sum(cell_count))
```

cells sequenced (each dataset may contain cells from several
ethnicities, as well as males or individuals of unknown gender, so we
do not know the actual number of cells available without downloading
files). Use `left_join` to identify the corresponding collections:

```{r}
## collections
left_join(
    african_american_female |> select(collection_id) |> distinct(),
    collections(db),
    by = "collection_id"
)
```

# Visualizing data in `cellxgene`

Discover files associated with our first selected dataset

```{r}
selected_files <-
    left_join(
        african_american_female |> select(dataset_id),
        files(db),
        by = "dataset_id"
    )
selected_files
```

The `filetype` column lists the type of each file. The cellxgene service
can be used to visualize *datasets* that have `CXG` files.

```{r, eval = FALSE}
selected_files |>
    filter(filetype == "CXG") |>
    slice(1) |> # visualize a single dataset
    datasets_visualize()
```

Visualization is an interactive process, so `datasets_visualize()`
will only open up to 5 browser tabs per call.

# File download and use

Datasets usually contain `CXG` (cellxgene visualization), `H5AD`
(files produced by the python AnnData module), and `Rds` (serialized
files produced by the _R_ Seurat package). There are no public parsers
for `CXG`, and the `Rds` files may be unreadable if the version of
Seurat used to create the file is different from the version used to
read the file. We therefore focus on the `H5AD` files. For
illustration, we download one of our selected files.

```{r}
local_file <-
    selected_files |>
    filter(
        dataset_id == "3de0ad6d-4378-4f62-b37b-ec0b75a50d94",
        filetype == "H5AD"
    ) |>
    files_download(dry.run = FALSE)
basename(local_file)
```

These are downloaded to a local cache (use the internal function
`cellxgenedp:::.cellxgenedb_cache_path()` for the location of the
cache), so the process is only time-consuming the first time.

`H5AD` files can be converted to _R_ / _Bioconductor_ objects using
the [zellkonverter][] package.

```{r}
h5ad <- readH5AD(local_file, reader = "R", use_hdf5 = TRUE)
h5ad
```

The `SingleCellExperiment` object is a matrix-like object with rows
corresponding to genes and columns to cells. Thus we can easily
explore the cells present in the data.

```{r}
h5ad |>
    colData(h5ad) |>
    as_tibble() |>
    count(sex, donor)
```

# Next steps

The [Orchestrating Single-Cell Analysis with Bioconductor][OSCA]
online resource provides an excellent introduction to analysis and
visualization of single-cell data in _R_ / _Bioconductor_. Extensive
opportunities for working with AnnData objects in _R_ but using the
native python interface are briefly described in, e.g., `?AnnData2SCE`
help page of [zellkonverter][].

[zellkonverter]: https://bioconductor.org/packages/zelkonverter
[OSCA]: https://bioconductor.org/books/OSCA

The [hca][] package provides programmatic access to the Human Cell
Atlas [data portal][HCAportal], allowing retrieval of primary as well
as derived single-cell data files.

[hca]: https://bioconductor.org/packages/hca
[HCAportal]: https://data.humancellatlas.org/explore

# Session info {.unnumbered}

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
